---
title: "RNAseq Analysis Example"
output: html_document
---
This is an introduction to RNAseq analysis for use at Software Carpentry bootcamps that have covered novice R. It involves reading in some count data from an RNAseq experiment, exploring the data and then analysis with the package DESeq2. 

```{r data_input}
## Filename with output from featureCounts
countfile <- "data/counts.txt"
## Read in the data
countdata <- read.table(countfile, header=TRUE, row.names=1)
head(countdata)
colnames(countdata)
class(countdata)
```
Have a look at the data. It contains genes (rows) with information about the gene positions in the first five columns and then information about the number of reads aligning to the gene in each experimental sample. We don't need the information on gene position, so we can remove it from the data frame.

```{r}

# Remove first five columns (chr, start, end, strand, length)
countdata <- countdata[ ,-(1:5)] 
head(countdata)
colnames(countdata)
```

We can rename the columns to something a bit more readable.
```{r}
## Manually
c("ctl1", "ctl2", "ctl3", "ctl4", "uvb1", "uvb2", "uvb3", "uvb4", "uvb5")
## Using paste and rep
?paste
paste("ctl", 1:4)
paste("ctl", 1:4, sep="")
c(paste("ctl", 1:4, sep=""), paste("uvb", 1:5, sep=""))
## Using gsub -- reproducible
?gsub
gsub(pattern=".fastq_tophat.accepted_hits.bam", replacement="", x=colnames(countdata))
colnames(countdata) <- gsub(pattern=".fastq_tophat.accepted_hits.bam", replacement="", x=colnames(countdata))
head(countdata)
```

DESeq2 analysis
---------

DESeq2 is an R package for analysis of RNAseq data. It is available from [Bioconductor](http://www.bioconductor.org/). [Explain packages and Bioconductor?]

```{r}
#install and have a break to check everyone is up to date?
source("http://bioconductor.org/biocLite.R")
biocLite("DESeq2")
#explain bioconductor?
library(DESeq2)

```
 It requires the count data to be in matrix form, and an additional dataframe describing the structure of the experiment.

```{r}
# Convert to matrix
class(countdata)
countdata <- as.matrix(countdata)
class(countdata)
head(countdata)

# construct coldata dataframe
#three replicates of control and UVB.
coldata <- data.frame(condition=c(rep("ctl", 3), rep("uvb",3)), row.names=colnames(countdata))
```

DESeq works on a particular type of object called a DESeqDataSet. 
```{r}
#introduce how DESeq2 works - type of object it works on etc
# instantiate the DESeqDataSet
dds <- DESeqDataSetFromMatrix(countData=countdata, colData=coldata, design=~condition)
dds
```

Run the DESeq pipeline on this object. [Describe pipeline steps?]
Get results and have a look at them
```{r}
dds <- DESeq(dds)

# Get differential expression results
res <- results(dds)
head(res)
table(res$padj<0.05)
## Order by adjusted p-value
res <- res[order(res$padj), ]
head(res)
```

Combine DEseq results with the original counts data. Write significant results to a file.
```{r}
resdata <- merge(as.data.frame(res), as.data.frame(counts(dds, normalized=TRUE)), by="row.names", sort=FALSE)
head(resdata)
names(resdata)[1] <- "GeneID"
head(resdata)

sig <- subset(resdata, padj<0.05)
write.table(sig, file="results/sig.txt", sep="\t") #tab delim data
```

Data Visualization
------------------

We can also do some exploratory plotting of the data.

```{r}
plotDispEsts(dds, main="Dispersion plot")
```

```{r}
## Regularized log transformation for clustering/heatmaps, etc
rld <- rlogTransformation(dds)
plotPCA(rld)

# Sample distance heatmap
head(assay(rld))
assay(rld)[1:5,1:5]
t(assay(rld))[1:5,1:5]
dist(t(assay(rld)))
as.matrix(dist(t(assay(rld))))
sampleDists <- as.matrix(dist(t(assay(rld))))
heatmap(sampleDists)
## better heatmap with gplots
install.packages("gplots")
library(gplots)
heatmap.2(sampleDists)
heatmap.2(sampleDists, col=colorpanel(64, "steelblue", "white"), key=FALSE, trace="none")
heatmap.2(sampleDists, col=colorpanel(64, "black", "white"), key=FALSE, trace="none")
heatmap.2(sampleDists, col=colorpanel(64, "red", "black", "green"), key=FALSE, trace="none")
heatmap.2(sampleDists, col=colorpanel(64, "red", "white", "blue"), key=FALSE, trace="none")

## Examine plot of p-values
hist(res$pvalue, breaks=50, col="grey")
```


```{r}
#These are the plots that are most recognisable from papers
# MA Plot
par(pch=16)
with(res, plot(baseMean, log2FoldChange, pch=20, cex=.5, log="x"))
with(subset(res, padj<.05), points(baseMean, log2FoldChange, col="red", pch=16))
## add points
install.packages("calibrate")
library(calibrate)
?textxy
res$Gene <- rownames(res)
with(subset(res, padj<.05), textxy(baseMean, log2FoldChange, labs=Gene, cex=1, col=2))

# Volcano plot
## Set point character
par(pch=16)
with(res, plot(log2FoldChange, -log10(pvalue), main="Volcano plot"))
with(subset(res, padj<.05 ), points(log2FoldChange, -log10(pvalue), col="red"))
with(subset(res, abs(log2FoldChange)>1), points(log2FoldChange, -log10(pvalue), col="orange"))
with(subset(res, padj<.05 & abs(log2FoldChange)>1), points(log2FoldChange, -log10(pvalue), col="green"))
## Add legend
legend("topleft", legend=c("FDR<0.05", "|LFC|>1", "both"), pch=16, col=c("red","orange","green"))
## Label points
with(subset(res, padj<.05 & abs(log2FoldChange)>1), textxy(log2FoldChange, -log10(pvalue), labs=Gene, cex=1))
```
